Certified incremental program transformations

### 2012-05-04 doc/papers/LFMTP2012/paper.tex
paper

\begin{abstract}
\end{abstract}

@@

\section{Introduction}

\subsection{The LF representation language}
\subsection{Storing computation results}
\subsection{Naming intermediate results}

\section{The calculus}

@@

The core calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\ ML
values --- binders and variables enabling $\lambda$-tree syntax (TODO
cite). Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical LF} (TODO cite) enriched with metavariables. We present
its typing algorithm relatively to a signature. It features an
explicit caching mechanism using metavariables.

@@

In this variant, only $\eta$-long $\beta$-normal forms have an
existence: canonicity for $\beta$-reduction is enforced in the syntax;
canonicity for $\eta$-expansion will be enforced by typing. Also,
application of a \emph{head} to a \emph{spine} (\ie\ a constant or
variable) is $n$-ary as opposed to the more standard binary
application: this eases reasoning about substitution and
$\eta$-expansion (TODO cite).

@@

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A\to B$ for $\prd x A B$ when $x\notin \FV{B}$,
$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list or
map notations for $\sigma$ and $\Gamma$, \ie\ write $\sigma(x) = M$ or
$\Gamma(x) = A$.

@@

We overload \emph{substitution} as a partial operation on terms, given
a parallel substitution $\sigma$. Since the syntax of object is not
stable by (textual) substitution, it is defined \emph{hereditarily}: it
is mutually recursive with a \emph{cut} function $\gcut M S$ which can
in turn trigger a chain of substitutions, leading to a canonical term,
a stuck substitution or divergence. % TODO vraiment?
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M N} S
  \\
  % TODO cas meta/S ????
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

We extend this operation trivially on families $\gsubst A \sigma$ and
kinds $\gsubst K \sigma$. We also allow to abusively write the
non-canonical application $\app M S$ as a shorthand for $\gcut M S$ if
it is defined.

Note that this notion of substitution presupposes $\eta$-long normal
forms: a non-empty spine cut against to another non-empty spine $\gcut
{\app H S} {S'}$ blocks the substitution.

% TODO pourquoi cette def?
% TODO terminaison??
% TODO stabilité de la eta?

@@

Typing an object while storing intermediate derivations requires
threading and maintaining a data structure of named partial
derivations (a repository) throughout the typing process. We present a
bidirectional typing algorithm updating an input repository.

@@

A \emph{signature} declares object and type constants to be used in
objects. To each object constant we assign an annotation $L$ setting
the slicing behaviour of the constant during type-checking: either
sliceable $\annsliceable$ (each application of this constant will have
its own name in the repository), non-sliceable $\annnonsliceable$ (we
don't create a name for applications of this constant), or defined
$\anndefined{F}$ (see \ref{sec:computational-calculus}).
\begin{align*}
  L &\gequal
  \annsliceable \gor
  \annnonsliceable \gor
  \anndefined{F}
  \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c :^L A} \gor
  \gcons \Sigma {\cst a :^L K}
  &\text{Signature}
\end{align*}
We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
suggesting the implementation of signatures as maps.

The algorithm is presented in Fig. \ref{fig:typing-algo} as a
syntax-directed inference system. All judgements are of the form $\gjo
U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
and $W$ the output. They are parameterized by a constant and implicit
signature $\Sigma$.

@@

  \begin{mathpar}
    \infer{
      \jstr{\Gamma\vdash\tp M A}{\Gamma'\vdash\tp {M'} {A'}}\sigma
      \and
      \text{$\meta X$ fresh in $\Delta$}
    }{
      \jp{(\Delta; \smeta Z \sigma)}\Gamma M A {(\mapadd\Delta{\meta
          X}{(\Gamma'\vdash M':A')}; \smeta X \sigma)}
    }
  \end{mathpar}

@@

@@

We now turn to the definition of our computation calculus on LF
objects. This language is untyped. % TODO??
\begin{align*}
  T &\gequal
  F \gor
  \matchin T \Gamma \casenobar {\app {\cst c} {\vec x}} T\ |\ \ldots\
  \case {\app {\cst c} {\vec x}} T
\end{align*}

@@

>>>>

%% Here is a problem
\begin{abstract}
  %% 4 sentences:
  %% State the problem
  %% Say why it’s an interesting problem
  %% Say what your solution achieves
  %% Say what follows from your solution
\end{abstract}

@@

\section*{Introduction}
%% It’s an interesting problem (1 page)

As programming languages develop more static analysis like rich type
systems, programs get safer as bugs are caught earlier.

%% Describe the problem (Use an example)
%% State your contributions (Bulleted list of contributions)

\section{Background}
%% It’s an unsolved problem (1 page)

\section{Here is my idea}
%% Here is my idea (2 pages)

\subsection{The \LF\ representation language}

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. The requirements are:
\begin{itemize}
\item conciseness of the certificates, since they are usually stored
  or transfered remotely; in particular,
\item a well-understood, short and independent verification algorithm;
  this is known as the \emph{de Bruijn principle}.
\item universality with respect to the particular problem: having a
  certificate language for each domain of application would just shift
  the problem of trust; having one universally accepted certificate
  language that can be tailored to many problems puts a much stronger
  trust on the unique verifier.
\end{itemize}

The Edimburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgements \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

This framework has been used to issue certificates for a number of
practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

In all these applications, the user feeds the generation program with
a query (the program to compile to machine-code, the access query, the
statement of a theorem \ldots) and gets in return a result (machine
code, yes/no\ldots) together with a certificate that the result
verifies the wanted property (semantics preservation, justification
for granting access, proof of the theorem\ldots). A potential client
can then verify independently that this result is adequate given the
query: it suffices to batch-verify the certificate.

\subsection{Proof-carrying static analysis}

\subsection{Storing computation results}
\subsection{Naming intermediate results}

\section{The calculus}
%% My idea works (details, data) (5 pages)

@@

The core calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. It features an explicit caching mechanism
using metavariables.

@@

In this variant, only $\eta$-long $\beta$-normal forms have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}.

@@

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A\to B$ for $\prd x A B$ when $\var x\notin \FV{B}$,
$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list, map
or set notations for $\sigma$ and $\Gamma$. The identity substitution
$\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var x}\ |\ \var x
\in\dom\Gamma\}$.

@@

We overload \emph{substitution} as a partial operation on objects, given
a parallel substitution $\sigma$. Since the syntax of object is not
stable by (textual) substitution, it is defined \emph{hereditarily}: it
is mutually recursive with a \emph{cut} function $\gcut M S$ which can
in turn trigger a chain of substitutions, leading to a canonical objects
or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ????
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

We extend this operation trivially on families $\gsubst A \sigma$ and
kinds $\gsubst K \sigma$. We also allow to abusively write
non-canonical applications $\app M S$ as a shorthand for $\gcut M S$
if they are defined.

Note that this notion of substitution differs from hereditary
substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
$\eta$-long normal forms: a non-empty spine cut against to another
non-empty spine $\gcut {\app H S} {S'}$ blocks the substitution
instead of reducing to $\app H {(\gconcat S S')}$ (an implicit
$\eta$-expansion). For example, the reduction $\gcut {\lam x \app
  {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because
the second $\var x$ in the left part is substituted by a function yet
is not itself a function.

% TODO pourquoi cette def?
% TODO terminaison??
% TODO stabilité de la eta?
% TODO comparer avec la def usuelle

@@

Typing an object while storing intermediate derivations requires
threading and maintaining a data structure of named partial
derivations (\emph{slices}) throughout the typing process. We present
a bidirectional typing algorithm updating an input repository.

A \emph{signature} declares object and type constants to be used in
objects. To each object constant we assign an annotation $L$ setting
the slicing behaviour of the constant during type-checking: either
sliceable $\annsliceable$ (each application of this constant will be
cached and given a name), non-sliceable $\annnonsliceable$ (we don't
create a name for applications of this constant), or defined
$\anndefined{T}$ (see \ref{sec:computational-calculus}).
\begin{align*}
  L &\gequal
  \annsliceable \gor
  \annnonsliceable \gor
  \anndefined{T}
  \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c :^L A} \gor
  \gcons \Sigma {\cst a :^L K}
  &\text{Signature}
\end{align*}
We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
suggesting the implementation of signatures as maps.

Our algorithm types exactly the well-typed objects of the usual
\LF, except that we thread a repository $\mr$ and store in it
all well-typed applicative subterms $F$ along with their classifier
$A$ and the environment in which they are well-typed.

% TODO parler de la redondance en LF normal

@@

We refer to cached subterms by metavariables. We force instantiation
of all variables in their environment by attaching a substitution to
them ($\smeta X \sigma$).

When checking large objects, the environment can grow fast. As an
optimization to reduce the size of the stored environments and
substitutions in slices $M$, we strengthen environments to their
minimum size (the free variables of $M$): We define the operation of
\emph{strengthening} from an environment $\Gamma$ and an object $M$
well-typed in $\Gamma$ to the minimal environment in which $M$ can
actually be typed, \ie\ containing only its free variables:
$$
\stren{\Gamma,M} = \{x : A\ |\ x\in \FV{M} \wedge \Gamma(x)=A\}
$$
% TODO lemme qui dit que c'est correct wrt variables libres des A et
% du type

The algorithm is presented in Fig. \ref{fig:typing-algo} as a
syntax-directed inference system. All judgements are of the form $\gjo
U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
and $W$ the output. They are parameterized by a constant and implicit
signature $\Sigma$, and all premisses are ordered in the algorithmical
order of computation.

@@

  \begin{mathpar}
    \infer{
      \stren{\Gamma, M}=\Gamma'
      \and
      \text{$\meta X$ fresh in $\Delta$}
    }{
      \jp{(\Delta; \smeta Z {\sigma})}\Gamma M A {(\mapadd\Delta{\meta
          X}{(\Gamma'\vdash M:A)}; \smeta X {\msubstid\Gamma})}
    }
  \end{mathpar}

@@

\begin{theorem}
  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
\end{theorem}

\begin{theorem}
  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutr
    M)} (\checkoutr A)$
\end{theorem}

% TODO example

@@

We now turn to the definition of our computation calculus. As we saw,
it is attached to special, \emph{defined} constants in the signature
and manipulates \LF\ objects. This language is un(i)typed, in that all
\LF\ objects have the same implicit type $\cst{obj}$. We have the
ability to construct objects out of other, and deconstruct them by
case analysis.
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma \casenobar {\app {\cst c} {\vec x}} U\ |\ \ldots\
  \case {\app {\cst c} {\vec x}} U
\end{align*}

@@

\section{Related work}
% 1-2 pages

\section*{Conclusion and future work}
% 0.5 pages
